SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-27 2017-Xudong-SMT/ActaInformatica.tex
reorganization 5.4, small bugs, and rewritten conclusion

For this example, the tool (using Algorithm 1) builds 258 open transitions, whereof 248
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 10 open
transitions, is shown in Figure~\ref{schema:resultOA1}.

@@

In the conception of the nanosatellite software \cite{CubETH-case-study}, the FailureTimer architecture was used to control several processes
communicating on the common bus. Timeout signals (``$ask$''), resets,
and the passing of time ($tick$) were synchronised in the global
system.

In Figure~\ref{schema:ArchFailure:BIP:Double} we show the combination
of two FailureTimer architectures, with their respective controlled
processes.
In Figure~\ref{schema:ArchFailure:Double:OA} we show the resulting
open automaton. It indicates that the two timers are forced to
be synchronous: each can only tick after the other has also started,
and then they will tick, and timeout, synchronously.

\TODO{Eric: try reorganise the figures. Add properties ?}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.9\columnwidth]{ActaXFIG/BIPspec-DoubleArchFailureTimer-v2-2}
  \caption{The BIP specification of two composed Failure Monitor architectures}
  \label{schema:ArchFailure:BIP:Double}
\end{figure}

@@

%% \begin{figure}[t]
%%   \centering
%%   \includegraphics[width=0.9\columnwidth]{ActaXFIG/DoubleFailureTimerPNET-v2-2}
%%   \caption{The open pNet encoding the Failure Monitor architecture (\ref{fig:schema:ArchFailure:BIP})}
%%   \label{fig:FailureTimer:pNet}
%% \end{figure}

%% \begin{figure}[t]
%%   \centering
%%   \includegraphics[width=0.9\columnwidth]{XFIG/DoubleTimerOA-v2-2}
%%   \caption{The open Automaton of the Failure Monitor Architecture}
%%   \label{fig:ArchFailure:OA}
%% \end{figure}

@@

All tests are run on HP xxx laptop under Windows 10 x64, with an Intel quad-Core i7-6600U
processor, and 32Gb of RAM. The software platform is VerCors\footnote{Available at \url{https://team.inria.fr/scale/software/vercors/}.}, running on top of
Eclipse Equinox, with Java 1.8.

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
\emph{open automaton}, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
In this paper, we used a running example, based on a BIP
architecture from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.  This encoding %\mdash which we intend to formalise and prove
% correct in a separate paper\mdash 
also opens the way for an extension
of BIP architectures with the transfer of data among variables of
different components.  Indeed, such data transfer can be easily
encoded using the predicates associated to synchronisation vectors in
open pNets.  The encoding of open transitions into SMTlib and the
availability of theories can guide the definition of such an extension.
%
%}
%
Our case-studies show that our encoding
successfully identifies the unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.

\TODO{add the novelties...}

%% Among the questions that have arised during this work, an important one is the
%% efficiency of Z3 for our needs. Z3 is famous for being very fast in
%% solving very large sets of assertions. %% , which could be important for
%% %% us
%% However, we encountered problems during the implementation of the
%% checking method. %% using Z3. 
%% For example, if the $\mathit{functions}$ and the
%% $\mathit{datatypes}$ part of the algebra presentation that we submit to the solver
%% are independent from each other, then checking the $\mathit{assertions}$ is 
%% simple. However, $\mathit{functions}$ on recursive $\mathit{datatypes}$ make it
%% more complex. Special rules might be defined by the user for the
%% induction.  We need more evidence that dealing with more complex
%% action algebras, which would involve axiomatizing their data structures
%% as Z3 theories, will allow us to decide validity of
%% predicates.
%% We may also try other automatic theorem provers, depending on the structure of the proofs we need.

%% \todoMargin{Change perspective: we are working on the formal extension of
%%   BIP architectures, and their semantics in terms of pNets. Then
%%   bisimulation and model-checking principles adapted to open pNet
%%   systems are under development}
%
Naturally, our next goals after the generation of the open automata
will be to model-check logical properties, and to check equivalences of pNets. While
model-checking open automata  seems easy to define, equivalence
checking is more challenging. 
In~\cite{henrio:Forte2016}, we
have already found the FH-bisimulation %inspired
%by~\cite{deSimone85},
to be a
suitable definition. But weak equivalences, or refinements, will
definitely be useful when comparing different pNets with
different structures. For bisimulation, we foresee that SMT methods
will be the basis for comparing open transitions. 
%However, the usual approach of computing bisimulation by partition
%% refinement does not work easily with symbolic transitions. In a first
%% step we may want instead to define and implement an algorithm for
%% checking that a given equivalence relation between open automata 
%% states is indeed a FH-Bisimulation. Finding automatically a
%% bisimulation relation or the most general bisimulation will be more
%% challenging.
%% It appears also that when comparing different pNets with
%% different structure, strong bisimulation will be too strong a notion,
%% and that we will have to define proper notions of weak equivalence or
%% refinements.

\emph{Scaling up.} One important motivation of this work is to
attack the complexity of verification of realistic systems by a
compositional and parametric approach. Still one may wonder if the price
for analysing our symbolic transitions will not make the approach too expensive
in terms of computing time. We tried a
slightly bigger example, assembling two Failure
controllers. In~\cite{Avocs-RR}, we show that Z3 can check the 
satisfiablility of  90K open transitions in a couple of minutes.

>>>>

For this example, the tool (using Algorithm 1) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.

@@

In the conception of the nanosatellite software \cite{CubETH-case-study}, the FailureTimer architecture was used to control several processes
communicating on the common bus. Timeout signals (``$ask$''), resets,
and the passing of time ($tick$) were synchronised in the global
system.
In this section, we use a combination of 2 FailureTimer architectures
to demonstrate how the encoding works, and as a slightly larger
example to show how the approach scales up.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\columnwidth]{ActaXFIG/BIPspec-DoubleArchFailureTimer-v2-2}
  \caption{The BIP specification of two composed Failure Monitor architectures}
  \label{schema:ArchFailure:BIP:Double}
\end{figure}

@@

In Figure~\ref{schema:ArchFailure:BIP:Double} we show the combination
of two FailureTimer architectures, with their respective controlled
processes.
In Figure~\ref{schema:ArchFailure:Double:OA} we show the resulting
open automaton. It indicates that the two timers are forced to
be synchronous: each can only tick after the other has also started,
and then they will tick, reset, and timeout, synchronously.

This is a very strong assumption, and we are working on a more
flexible specification (at BIP level) to release this hypothesis while
keeping the safety properties of the composition.

@@

All tests are run on HP EliteBook 840 laptop under Windows 10 x64, with an Intel quad-Core i7-6600U
processor, and 32Gb of RAM. The software platform is VerCors\footnote{Available at \url{https://team.inria.fr/scale/software/vercors/}.}, running on top of
Eclipse Equinox, with Java 1.8.

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
\emph{open automaton}, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
We have used a running example based on two possible
encodings of the Enable operator of Lotos. The automata computed with
the algorithm can be used to prove equational properties of the
operator.

Then we define a ``smart'' version of the algorithm that explores, as
much as possible, only the reachable part of the open automaton. The
handling of assignments makes the reachable computation more
challenging. We show that the result is equivalent to the original
algorithm.

We validate the approach on two larger examples, 
based on a BIP architecture taken from an earlier nanosatellite case
study~\cite{CubETH-case-study}. We also show how such architectures
can be composed, and how our approach scales.
This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.
We also compare the practical performances of the 2 algorithms and
show the benefit gained using the smart approach.

Naturally, our next goals after the generation of the open automata
will be to model-check logical properties, and to check equivalences
of pNets. While model-checking open automata seems easy to define,
equivalence checking is more challenging. 
In~\cite{henrio:Forte2016}, we
have already found the FH-bisimulation 
to be a suitable definition. But weak equivalences, or refinements, will
definitely be useful when comparing different pNets with
different structures. For both model-checking and bisimulation, SMT methods
will be the basis for comparing open transitions.

%% \emph{Scaling up.} One important motivation of this work is to
%% attack the complexity of verification of realistic systems by a
%% compositional and parametric approach. Still one may wonder if the price
%% for analysing our symbolic transitions will not make the approach too expensive
%% in terms of computing time. We tried a
%% slightly bigger example, assembling two Failure
%% controllers. In~\cite{Avocs-RR}, we show that Z3 can check the 
%% satisfiablility of  90K open transitions in a couple of minutes.
